The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with {\em Gaia} Data and Beyond

### 2019-04-26 paper/ms.tex
Fix epicyclic section

Under this approximation, we can write down the cylindrical components
of the orbits as,
\beq\label{eq:orbits_epi}
\begin{split}
\phi(t) &= \Omega_c t \\
R(t) &= R_g + A_R \sin{(\nu t + \beta)} \\
z(t) &= A_z \sin{(\kappa t + \alpha)}
\text{,}
\end{split}
\eeq
where $\kappa$ and $\nu$ are the vertical and radial (or epicyclic)
frequencies, $\Omega_c \equiv v_c/R_g$ the rotation frequency, $A_z$ and $A_R$ are the amplitudes of the stellar motion in the
vertical and radial coordinates, and $\alpha$ and $\beta$ are arbitrary
constants. Similarly, the velocities of the orbit are given by
\beq\label{eq:orbits_vel_epi}
\begin{split}
v_{\phi}(t) &= v_c \\
v_R(t) &= A_R \nu \cos{(\nu t + \beta)} \\
v_z(t) &= A_z \kappa \cos{(\kappa t + \alpha)}
\text{.}
\end{split}
\eeq

In this case, we have that for $J_{\phi}$ \citep[][Section~3.5.3b]{2008gady.book.....B}, 
\beq\label{eq:Jphi_epi}
J_{\phi} = R_g v_c\text{,}
\eeq
and for $J_R$ and $J_z$,
\beq\label{eq:JR_Jz_epi}
\begin{split}
J_R &= \frac{E_R}{\kappa} \\
J_z &= \frac{E_z}{\nu} \text{,}
\end{split}
\eeq
where $E_R$ and $E_z$ are the energy per unit mass in the radial and vertical
coordinates, respectively. Since $a_R = -\kappa^2(R-R_g)$, and $a_z = - \nu^2
z$, we have that,
\beq\label{eq:JR_Jz_epi_energy}
\begin{split}
J_R &= \frac{\frac{1}{2}v_R^2 + \kappa^2 (R-R_g)^2}{\kappa} \\
J_z &= \frac{\frac{1}{2}v_z^2 + \nu^2 z^2}{\nu}
\end{split}
\eeq

While the formula for $J_{\phi}$ is phase independent, $J_R$ and $J_z$ are
both phase dependent via Equation~\eqref{eq:orbits_epi}. We consider two cases. First, at the point of maximum
amplitude or when $(v_R, v_z) = (0, 0)$ and second, at the point of maximum
velocity $(R, z) = (R_g, 0)$. For $J_R$ we have that,
\beq\label{eq:JR_epi_final}
    \begin{aligned}
        J_R &= \kappa A_R^2 && \text{if $v_R = 0$} \\
    J_R &= \frac{v_{R,\text{max}}^2}{2\kappa^2} && \text{if $R=R_g$} \text{.}
    \end{aligned}
\eeq
For $J_z$, we have that,
\beq\label{eq:Jz_epi_final}
    \begin{aligned}
        J_z &= \kappa A_z^2 && \text{if $v_z = 0$} \\
        J_z &= \frac{v_{z,\text{max}}^2}{2\kappa^2} && \text{if $z=0$} \text{.}
    \end{aligned}
\eeq

Now assume that the coordinates $(R, z, v_{\phi}, v_R, v_z)$ are offset by
$(\Delta R, \Delta z, \Delta v_{\phi}, \Delta v_R, \Delta v_z)$. Then, the
induced error in $J_{\phi}$ is, regardless of phase,
\beq\label{eq:induced_Jphi}
\frac{\Delta J_{\phi}}{J_{\phi}} = \frac{\Delta R}{R_g}\text{.}
\eeq
For $J_R$, the induced error is,
\beq\label{eq:induced_JR}
\begin{aligned}
    \frac{\Delta J_R}{J_R} &= \frac{2\Delta R}{A_R} + 
        \left(\frac{\Delta v_R}{v_{R,\text{max}}}\right)^2 && \text{if $v_R = 0$} \\
    \frac{\Delta J_R}{J_R} &= \frac{2\Delta v_R}{v_{R,\text{max}}} + 
        \left(\frac{\Delta R}{A_R}\right)^2  && \text{if $R=R_g$} \text{.}
\end{aligned}
\eeq
For $J_{z}$, the induced error is,
\beq\label{eq:induced_Jz}
\begin{aligned}
    \frac{\Delta J_z}{J_z} &= \frac{2\Delta z}{A_z} + 
        \left(\frac{\Delta v_z}{v_{z,\text{max}}}\right)^2 && \text{if $v_z = 0$} \\
    \frac{\Delta J_z}{J_z} &= \frac{2\Delta v_z}{v_{z,\text{max}}} + 
        \left(\frac{\Delta z}{A_z}\right)^2  && \text{if $z=0$} \text{.}
\end{aligned}
\eeq

>>>>

Under this approximation, we can write down the cylindrical components
of the orbits as,
\beq\label{eq:orbits_epi}
\begin{split}
\phi(t) &= \Omega_c t \\
R(t) &= R_g + A_R \sin{(\kappa t + \alpha)} \\
z(t) &= A_z \sin{(\nu t + \beta)}
\text{,}
\end{split}
\eeq
where $\kappa$ and $\nu$ are the radial and vertical
frequencies, $\Omega_c \equiv v_c/R_g$ the rotation frequency, $A_z$ and $A_R$ are the amplitudes of the stellar motion in the
vertical and radial coordinates, and $\alpha$ and $\beta$ are arbitrary
constants. Similarly, the velocities of the orbit are given by
\beq\label{eq:orbits_vel_epi}
\begin{split}
v_{\phi}(t) &= v_c \\
v_R(t) &= \kappa A_R \cos{(\kappa t + \alpha)} \\
v_z(t) &= \nu A_z \cos{(\nu t + \beta)}
\text{.}
\end{split}
\eeq

In this case, we have that for $J_{\phi}$ \citep[][Section~3.5.3b]{2008gady.book.....B}, 
\beq\label{eq:Jphi_epi}
J_{\phi} = R_g v_c\text{,}
\eeq
and for $J_R$ and $J_z$,
\beq\label{eq:JR_Jz_epi}
\begin{split}
J_R &= \frac{E_R}{\kappa} \\
J_z &= \frac{E_z}{\nu} \text{,}
\end{split}
\eeq
where $E_R$ and $E_z$ are the energy per unit mass in the radial and vertical
coordinates, respectively. Therefore,
\beq\label{eq:JR_Jz_epi_energy}
\begin{split}
J_R &= \frac{v_R^2 + \kappa^2 (R-R_g)^2}{2\kappa} \\
J_z &= \frac{v_z^2 + \nu^2 z^2}{2\nu}\text{.}
\end{split}
\eeq
We can simplify this and write,
\beq\label{eq:JR_Jz_epi_final}
\begin{split}
J_R &= \frac{\kappa A_R^2}{2} = \frac{v_{R,\text{max}}^2}{2\kappa} \\
J_z &= \frac{\nu A_z^2}{2} = \frac{v_{z,\text{max}}^2}{2\nu}\text{,}
\end{split}
\eeq
where the last equality comes from the fact that $v_{R,\text{max}} = \kappa
A_R$ and similarly for $z$.

Notice that while the value for each of $J_{\phi}$, $J_R$, and $J_z$ is phase
independent, the contribution from the kinetic and potential terms in
Equation~\eqref{eq:JR_Jz_epi_energy} is phase dependent. Now assume that the
coordinates $(R, z, v_{\phi}, v_R, v_z)$ are offset by $(\Delta R, \Delta z,
\Delta v_{\phi}, \Delta v_R, \Delta v_z)$. We can then apply the standard
propagation of errors formula to Equation~\eqref{eq:JR_Jz_epi_energy}
determine the error in each of the actions. For $J_{\phi}$ we have that,
\beq\label{eq:induced_Jphi}
\frac{\Delta J_{\phi}}{J_{\phi}} = \frac{\Delta R}{R_g} + \frac{\Delta v_{\phi}}{v_c}\text{.}
\eeq
For $J_R$, the induced error is,
\beq\label{eq:induced_JR}
\frac{\Delta J_R}{J_R} = \frac{2(R-R_g)}{A_R^2}\Delta R
                         + \frac{2v_R}{v_{R,\text{max}}^2} \Delta v_R \text{.}
\eeq
For $J_z$, the induced error is,
\beq\label{eq:induced_Jz}
\frac{\Delta J_z}{J_z} = \frac{2z}{A_z^2}\Delta z
                         + \frac{2v_z}{v_{z,\text{max}}^2} \Delta v_z \text{.}
\eeq
We have ignored second order contributions.
